Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2025-05-02 analysis/paper/article.qmd
replace plot of changing requirements over time to match reviewer's suggestion

# save these three plots as one set
library(cowplot)
p <- 
plot_grid(
 fig_prop_by_job_title_per_year + theme(plot.margin = unit(c(0, 0, 0, 0), "cm")),
  fig_ratio_tt_2_ntt_jobs_per_year + theme(plot.margin = unit(c(0, 0, 0, 0), "cm")),

@@

# combine list into one big data frame
edits_for_each_year_tbl <-
bind_rows(edits_urls_for_each_year_lst_tbl_with_comments,
          .id = "year_ad_posted") %>%
  mutate(
    edit_size_num = parse_number(str_replace(edit_size, "−", "-")),
    edit_time = parse_date_time(edit_date,
                                 orders = "HM, d B Y"),
    # round down to first day of month
    edit_month = floor_date(edit_time,
                            unit = "month"),
    edit_year = floor_date(edit_month,
                           unit = "year"),
    edit_year_only = year(edit_month),
    job_market_year = parse_number(str_sub(year_ad_posted,
                              -4L))
  ) %>%
  # delete edits that were made after the job market year
  filter(edit_year_only == job_market_year | edit_year == (job_market_year -1))

@@

# Number of edits per year
p_edits_per_year <-
ggplot(edits_for_each_year_tbl) +
  aes(as_date(edit_year)) +
  geom_bar() +
  theme_minimal(base_size = small_base_size) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Number of edits") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

# Number of unique editors per year
p_editors_per_year <-
edits_for_each_year_tbl %>%
  group_by(edit_year) %>%
  summarise(n_distinct_editors = n_distinct(edit_name)) %>%
ggplot() +
  aes(as_date(edit_year),
      n_distinct_editors) +
  geom_col() +
  theme_minimal(base_size = small_base_size) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Number of editors\n(distinct usernames or IP addresses)") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

@@

# life span of an editor, from first edit to last edit
edits_for_each_year_tbl_span <-
edits_for_each_year_tbl %>%
  group_by(edit_name) %>%
  summarise(first_edit = min(edit_month),
            last_edit =  max(edit_month),
            edit_span = last_edit - first_edit,
            edit_span_years = as.numeric(edit_span,
                                         units = "days") / (365.25)) %>%
  mutate(first_edit_year = year(first_edit),
         last_edit_year =  year(last_edit))

@@

# for those editors active for more than three months:
p_editor_life_distr_year <-
edits_for_each_year_tbl_span %>%
 filter(edit_span_years > 0.25) %>%
ggplot() +
  aes(as_date(first_edit_year),
      reorder(edit_name, first_edit_year)) +
  geom_segment(aes(x =   first_edit,
                   xend = last_edit),
               linewidth = 1,
               lineend = "butt") +
  theme_minimal(base_size = small_base_size) +
  labs(x = "", y = "Individual editors\n(usernames and IP addresses redacted)") +
  theme(axis.text.y=element_blank())

# size of edits per year
p_edit_size_per_year <-
edits_for_each_year_tbl %>%
  group_by(edit_year) %>%
  summarise(edit_size_num = sum(edit_size_num)) %>%
  ggplot() +
  aes(as_date(edit_year),
      edit_size_num) +
  geom_col() +
  theme_minimal(base_size = small_base_size) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Sum of edits (Bytes)") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

# typical edit size per year
p_median_edit_size_per_year <-
edits_for_each_year_tbl %>%
  group_by(edit_year) %>%
  # median, because sometimes there are 1-2 very large
  # page reformatting edits
  summarise(edit_size_num = median(edit_size_num)) %>%
  ggplot() +
  aes(as_date(edit_year),
      edit_size_num) +
  geom_col() +
  theme_minimal(base_size = small_base_size) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Median edit size (Bytes)") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

@@

# edits per section for each year
library(ggbeeswarm)
p_edit_number_per_entry_per_year <-
ggplot(section_edit_tally_tbl) +
  aes(as_date(edit_year),
      group = as_date(edit_year),
      n) +
  geom_boxplot() +
  theme_minimal(base_size = small_base_size) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Number of edits per job ad") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))
```

@@

how_many_people_use_the_wiki <- 
tidy_table_to_plot %>% 
  filter(status != "How many people use the wiki?") %>% 
  mutate(status = str_wrap(status, 26)) %>% 
ggplot() +
  aes(year,
      count,
      fill = status) +
  geom_col() +
  theme_minimal(base_size = small_base_size) +
  xlab("") +
  scale_fill_manual(values = the_thing_palette) +
  theme(legend.position = c(.7,.8),
        legend.key.size = unit(0.25,"line"),
        legend.text = element_text(margin = margin(r = 10, 
                                                   t = 2,
                                                   b = 2,
                                                   unit = "pt"))) +
  guides(fill=guide_legend(nrow=3,
                           byrow=TRUE),
         )
```

```{r}
#| label: fig-panel-wiki-characteristics
#| fig-cap: "A: Total number of edits to each Wiki page (additions and deletions) per year. B: Number of unique editors, as represented by distinct usernames or IP addresses (recorded for editors without usernames), active per year. C: Active periods of individual editors, the black lines connect the date of the first and last edit of all editors that were active for three months or more. D: Typical size of each edit per year, either addition or removal of text, one word is roughly 4-6 Bytes. E: Distribution of the numbers of edits per job ad for each year. F: Histogram showing that the majority of editors are only active on the site for less than one year. G: Breakdown of self-reported Wiki page user categories for each year."
#| fig-height: 5

@@

Because there is no single editorial authority, collaboratively-edited resources such as the Academic Jobs Wiki can be highly variable in the amount and type of activity over time. It is important to characterise this activity to assess the reliability of the content is as a source of information about the job market. We can get some insights into the dynamics of editing activity from the MediaWiki technology used by Academic Jobs Wiki that creates a public record of every act of adding or removing text to any page in the Wiki [@barrett2008mediawiki]. @fig-panel-wiki-characteristics shows the temporal trends in editing of the Academic Jobs Wiki. The number of edits and editors roughly halved from 2013 to 2017, and while the number of edits remained relatively steady after 2017, the number of editors continued to gradually decline. The vast majority of editors are active on the site for less than a year (n = r` edits_for_each_year_tbl_span_tbl_less_than_one_year`, `r edits_for_each_year_tbl_span_tbl_less_than_one_year_perc`%), perhaps reflecting a single season of job-searching. Editors with a longer history likely come from two groups, job seekers participating in multiple seasons of job applications, and faculty on search committees posting multiple job ads.

@@

```{r}
#| label: fig-show-basic-plots
#| fig-cap: "A: total number of job ads posted to the Academic Jobs Wiki for Archaeology in each year, with coloured sections showing the proportion of jobs by title and rank. B: Ratio of tenure-track to non-tenure-track positions over time. The red line indicates a 1:1 ratio of tenure-track to non-tenure-track positions, bars taller than that line indicate more tenure-track than non-tenure-track positions in that year"

@@

p_req <- 
ggplot(jobdata_requirements_means) +
  aes(year_ad_posted_break, 
      mean_n,
      group = name) +
  geom_smooth(linewidth = 2,
              colour = "black") +
  geom_jitter(data = jobdata_requirements,
             aes(year_ad_posted_break, 
                 value),
             alpha = 0.1,
             height = 0.2,
             width =  0.1) +
  facet_wrap(~name,
             scales = "free_y",
             nrow = 2) +
  xlab("") +
  ylab("Number requested in job ad") +
  scale_y_continuous(breaks = integer_breaks()) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(size = 8),
        strip.text = element_text( size = 16))

@@

```{r}
#| label: fig-requirements-over-time
#| fig-cap: "Changing requirements in job ads over time. Each point represents an individual job ad. The points have been jittered to avoid overlap. The black line is a locally weighted regression line to identify temporal trends. The grey region is the 95% confidence interval for the regression line."

@@

Over our ten year study period there have been substantial changes in the instructions to applicants in terms of the type and number of documents that are requested by the search committees [@fig-requirements-over-time]. Requests for cover letter and CV decline in more recent years, perhaps reflecting their status as a generic and expected component to submit without needing to be specifically requested. The requirement for a diversity statement is rare until 2019-2020, peaks around 2020-2021, then decreases towards the present. Requests for names of recommenders (either zero or three names, rarely only two names) reaches a maximum during 2019-2020, then also decline for the remainder of the study period. The requirement for a research statement and teaching statement increases after 2015-2016, and becomes more frequent in job ads in more recent years. Requests for course descriptions, syllabus samples, teaching evaluations, transcripts and writing samples are consistently low over time (not shown here).

>>>>

# save these three plots as one set
library(cowplot)
p <- 
plot_grid(
 fig_prop_by_job_title_per_year_us + theme(plot.margin = unit(c(0, 0, 0, 0), "cm")),
  fig_ratio_tt_2_ntt_jobs_per_year + theme(plot.margin = unit(c(0, 0, 0, 0), "cm")),

@@

# combine list into one big data frame
edits_for_each_year_tbl <-
bind_rows(edits_urls_for_each_year_lst_tbl_with_comments,
          .id = "year_ad_posted") %>%
  mutate(
    edit_size_num = parse_number(str_replace(edit_size, "−", "-")),
    edit_time = parse_date_time(edit_date,
                                 orders = "HM, d B Y"),
    # round down to first day of month
    edit_month = floor_date(edit_time,
                            unit = "month"),
    edit_year = floor_date(edit_month,
                           unit = "year"),
    edit_year_only = year(edit_month),
    job_market_year = parse_number(str_sub(year_ad_posted,
                              -4L))
  ) # %>%
  # delete edits that were made after the job market year
  # filter(edit_year_only == job_market_year | edit_year == (job_market_year -1))

@@

# Number of edits per year
p_edits_per_year <-
  edits_for_each_year_tbl %>% 
  filter(between(edit_year,
                 as.POSIXct("2012-01-01 0:00:00"), 
                 as.POSIXct("2023-01-01 0:00:00"))) %>% 
ggplot() +
  aes(as_date(edit_year)) +
  geom_bar() +
  theme_minimal(base_size = small_base_size) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Number of edits") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

# Number of unique editors per year
p_editors_per_year <-
edits_for_each_year_tbl %>%
    filter(between(edit_year,
                 as.POSIXct("2012-01-01 0:00:00"), 
                 as.POSIXct("2023-01-01 0:00:00"))) %>% 
  group_by(edit_year) %>%
  summarise(n_distinct_editors = n_distinct(edit_name)) %>%
ggplot() +
  aes(as_date(edit_year),
      n_distinct_editors) +
  geom_col() +
  theme_minimal(base_size = small_base_size) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Number of editors\n(distinct usernames or IP addresses)") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

@@

# life span of an editor, from first edit to last edit
edits_for_each_year_tbl_span <-
edits_for_each_year_tbl %>%
    filter(between(edit_year,
                 as.POSIXct("2012-01-01 0:00:00"), 
                 as.POSIXct("2023-01-01 0:00:00"))) %>% 
  group_by(edit_name) %>%
  summarise(first_edit = min(edit_month),
            last_edit =  max(edit_month),
            edit_span = last_edit - first_edit,
            edit_span_years = as.numeric(edit_span,
                                         units = "days") / (365.25)) %>%
  mutate(first_edit_year = year(first_edit),
         last_edit_year =  year(last_edit))

@@

# for those editors active for more than three months:
p_editor_life_distr_year <-
edits_for_each_year_tbl_span %>%
 filter(edit_span_years > 0.25) %>%
ggplot() +
  aes(as_date(first_edit_year),
      reorder(edit_name, first_edit_year)) +
  geom_segment(aes(x =   first_edit,
                   xend = last_edit),
               linewidth = 0.1,
               lineend = "butt") +
  theme_minimal(base_size = small_base_size) +
  labs(x = "", y = "Individual editors\n(usernames and IP addresses redacted)") +
  theme(axis.text.y=element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),)

# size of edits per year
p_edit_size_per_year <-
edits_for_each_year_tbl %>%
    filter(between(edit_year,
                 as.POSIXct("2012-01-01 0:00:00"), 
                 as.POSIXct("2023-01-01 0:00:00"))) %>% 
  group_by(edit_year) %>%
  summarise(edit_size_num = sum(edit_size_num)) %>%
  ggplot() +
  aes(as_date(edit_year),
      edit_size_num) +
  geom_col() +
  theme_minimal(base_size = small_base_size) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Sum of edits (Bytes)") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

# typical edit size per year
p_median_edit_size_per_year <-
edits_for_each_year_tbl %>%
    filter(between(edit_year,
                 as.POSIXct("2012-01-01 0:00:00"), 
                 as.POSIXct("2023-01-01 0:00:00"))) %>% 
  group_by(edit_year) %>%
  # median, because sometimes there are 1-2 very large
  # page reformatting edits
  summarise(edit_size_num = median(edit_size_num)) %>%
  ggplot() +
  aes(as_date(edit_year),
      edit_size_num) +
  geom_col() +
  theme_minimal(base_size = small_base_size) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Median edit size (Bytes)") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))

@@

# edits per section for each year
library(ggbeeswarm)
p_edit_number_per_entry_per_year <-
section_edit_tally_tbl %>% 
    filter(between(edit_year,
                 as.POSIXct("2012-01-01 0:00:00"), 
                 as.POSIXct("2023-01-01 0:00:00"))) %>% 
  ggplot() + 
  aes(as_date(edit_year),
      group = as_date(edit_year),
      n) +
  geom_boxplot(outlier.size = 0.5) +
  theme_minimal(base_size = small_base_size) +
  scale_x_date(date_breaks = "1 year",
               date_labels = "%Y") +
  labs(x = "", y = "Number of edits per job ad") +
  theme(axis.text.x = element_text(angle = 90,
                                   vjust = 0.5))
```

@@

how_many_people_use_the_wiki <- 
tidy_table_to_plot %>% 
  filter(status != "How many people use the wiki?") %>% 
  mutate(status = str_wrap(status, 26)) %>% 
ggplot() +
  aes(year,
      count,
      fill = status) +
  geom_col() +
  theme_minimal(base_size = small_base_size) +
  xlab("") +
  scale_fill_manual(values = the_thing_palette) +
  theme(legend.position = c(.7,.8),
        legend.key.size = unit(0.25,"line"),
        legend.text = element_text(margin = margin(r = 10, 
                                                   l = 2,
                                                   t = 2,
                                                   b = 2,
                                                   unit = "pt"))) +
  guides(fill=guide_legend(nrow=3,
                           byrow=TRUE),
         )
```

```{r}
#| label: fig-panel-wiki-characteristics
#| fig-cap: "A: Total number of edits to each Wiki page (additions and deletions) per year. B: Number of unique editors, as represented by distinct usernames or IP addresses (recorded for editors without usernames), active per year. C: Active periods of individual editors, the black lines connect the date of the first and last edit of all editors that were active for three months or more. D: Typical size of each edit per year, either addition or removal of text, one word is roughly 4-6 Bytes. E: Distribution of the numbers of edits per job ad for each year. F: Histogram showing that the majority of editors are only active on the site for less than one year. G: Breakdown of self-reported Wiki user categories for each year."
#| fig-height: 5

@@

```{r}
# geolocate users by IP address
# devtools::install_github("ironholds/rgeolocate")

library(rgeolocate)
file <- system.file("extdata","GeoLite2-Country.mmdb", package = "rgeolocate")
results <- maxmind(unique(edits_for_each_year_tbl$edit_name),
                   file,
                   "country_code")

n_all_user_ids <- length(unique(edits_for_each_year_tbl$edit_name))

results_ip_addresses <-
  results %>%
  drop_na()

n_all_user_ips <-  nrow(results_ip_addresses)

# percentage of users that we can geolocate
per_users_we_can_geolocate <- 
  round(n_all_user_ips / n_all_user_ids * 100, 0)

# tally and percentage
results_ip_addresses_tbl <- 
results_ip_addresses %>%
  group_by(country_code) %>%
  tally(sort = TRUE) %>%
  mutate(perc = n / sum(n) * 100)

results_ip_addresses_tbl_us_scl <- 
  results_ip_addresses_tbl %>% 
  filter(country_code == "US") %>% 
  pull(perc) %>% 
  round(0)

Because there is no single editorial authority, collaboratively-edited resources such as the Academic Jobs Wiki can be highly variable in the amount and type of activity over time. It is important to characterise this activity to assess the reliability of the content is as a source of information about the job market. We can get some insights into the dynamics of editing activity from the MediaWiki technology used by Academic Jobs Wiki that creates a public record of every act of adding or removing text to any page in the Wiki [@barrett2008mediawiki]. @fig-panel-wiki-characteristics shows the temporal trends in editing of the Academic Jobs Wiki. Every edit includes the author's identity recorded as either an IP address (a numerical address that includes information about the location of the user's computer) or a username. Of the `r n_all_user_ids` unique editors in our sample, `r per_users_we_can_geolocate`% used IP addresses rather than usernames, and geolocation analysis shows that `r results_ip_addresses_tbl_us_scl`% of IP addresses were based in the US, indicating the Wiki is primarily used by archaeologists in the US. The number of edits and editors roughly halved from 2013 to 2017, and continued to gradually decline. Despite this trend, the size of individual edits slightly increases over time, and the number of edits per job ad is relatively constant over time. The majority of editors are active on the site for less than a year (n = r` edits_for_each_year_tbl_span_tbl_less_than_one_year`, `r edits_for_each_year_tbl_span_tbl_less_than_one_year_perc`%), perhaps reflecting a single season of job-searching. Editors with a longer history likely come from two groups, job seekers participating in multiple seasons of job applications, and faculty on search committees posting multiple job ads.

@@

```{r}
#| label: fig-show-basic-plots
#| fig-cap: "A: total number of job ads from US institutions posted to the Academic Jobs Wiki for Archaeology in each year, with coloured sections showing the proportion of jobs by title and rank. B: Ratio of tenure-track to non-tenure-track positions over time. The red line indicates a 1:1 ratio of tenure-track to non-tenure-track positions, bars taller than that line indicate more tenure-track than non-tenure-track positions in that year"

@@

p_req <- 
jobdata_requirements %>%
  mutate(name = case_when(
    name == "cover letter" ~ "Cover letter",
    name == "cv" ~ "CV",
    name == "research statement" ~ "Research statement",
    name == "teaching statement" ~ "Teaching statement",
    name == "diversity statement" ~ "Diversity statement",
    name == "names of recommenders" ~ "Names of recommenders")) %>%
  ggplot() +
  aes(year_ad_posted_break,
      fill = factor(value),
      group = factor(value)) +
  geom_bar(stat="count",
           position = "fill") +
  facet_wrap(~name,
             scales = "free_y",
             nrow = 2) +
  xlab("") +
  ylab("Proportion of all job ads in a year") +
  scale_y_continuous(breaks = integer_breaks()) +
  theme_minimal(base_size = base_size) +
  scale_fill_brewer(palette = "Dark2") +
  guides(fill = guide_legend("Count")) +
  theme(axis.text.x = element_text(size = 8),
        strip.text = element_text( size = 14))

@@

```{r}
#| label: fig-requirements-over-time
#| fig-cap: "Changing requirements in job ads over time. 'Count' refers to the number of each item required."

@@

Over our ten year study period there have been substantial changes in the instructions to applicants in terms of the type and number of documents that are requested by the search committees [@fig-requirements-over-time]. Requests for cover letter and CV decline slightly in more recent years, perhaps reflecting their status as a generic and expected component to submit without needing to be specifically requested. The requirement for a diversity statement is rare until 2019-2020, peaks around 2020-2021, then decreases towards the present. Requests for names of recommenders (either zero or three names, rarely only two names) reaches a maximum during 2019-2020, then also decline for the remainder of the study period. The requirement for a research statement and teaching statement increases after 2015-2016, and becomes more frequent in job ads in more recent years. Requests for course descriptions, syllabus samples, teaching evaluations, transcripts and writing samples are consistently low over time (not shown here).
